### 2022-02-18 pre/pre.tex
more

Atomic formulas of propositional logic are propositional variables
$\cst{p}, \cst{q}, \ldots$ and constants $\top$ and $\bot$. More complex
formulas are built inductively using logical connectives $\inot, \iand, \ior,
\iimplies, \allowbreak \iequiv$: if $\phi$ and $\psi$ are formulas, then $\inot \phi, \phi \iand
\psi, \allowbreak \phi \ior \psi, \phi \iimplies \psi, \phi \iequiv \psi$ are formulas as well.

To interpret the formulas, propositional variables are assigned 0 (false) or 1
(true), constants $\top$ and $\bot$ are assigned 1 and 0 respectively and the
formula is interpreted using the rules for each connective
\cite[Sect.~1.4]{hr-00-logic-in-cs}. As there are finitely many propositional
variables in a formula, trying all (finitely many) possible variable assignments
describes an algorithm to decide satisfiability (existence of satisfiable
assignment for a formula) or validity (if all assignments are satisfy the formula).

@@

We distinguish a set of base types $T$ which is required to have the Boolean
type $o$, and a set of symbols $\Sigma$. To each symbol $\cst{f} \in \Sigma$ a
tuple $(\tau_1, \ldots, \tau_{n}, \tau), n \geq 0$ of types is assigned, written
$\cst{f} : (\tau_1, \ldots, \tau_{n}) \rightarrow \tau$. We say that $(\tau_1,
\ldots, \tau_{n})$ are argument types, $\tau$ is the return type, and that $n$
is the arity of the symbol $\cst{f}$. If return type of $\cst{f}$ is $o$ we call
$\cst{f}$ a predicate symbol, otherwise we call it a function symbol. Argument types may not be Boolean.

Terms are the basic building blocks of first-order logic. Variables
$x,y,\ldots$, assigned types $\tau \in T$, are terms. If $t_1,\ldots,t_n$ are
terms of types $\tau_1, \ldots, \tau_n$, respectively, and $\cst{f} : (\tau_1,
\ldots, \tau_{n}) \rightarrow \tau$, then $\cst{f}(t_1, \ldots, t_n)$ is a term
of type $\tau$. If $n=0$, we drop parentheses and write $\cst{f}$. We also
abbreviate $\cst{f}(t_1, \ldots, t_n)$ to $\cst{f}(\tuplen{t})$. Using similar
notation, we abbreviate a tuple of terms $(t_1, \ldots, t_n)$ as $(\tuplen{t})$
or simply $\tuplen{t}$.

Terms are used to build atoms. Atom is either a term $t$ of type $o$ or an
equation $s \eq t$ where terms $t$ and and $s$ are of the same type. Atoms are
combined using logical connectives to build formulas just like in propositional
logic. Additionally, first-order formulas are built using quantifiers: if $\phi$
is a formula and $x$ is a variable then $\iforall x.\, \phi$, as well as
$\iexists x.\, \phi$ are formulas. The first formula requires that $\phi$ holds
for any value of $x$, while the second one requires that there is some $x$ for
which $\phi$ holds. As soon as there is a single functional symbol with arity
greater than 0, there are infinitely many terms that can be substituted for a
free variable. Furthermore, in first-order logic to determine the truth value of
the formula, it does not suffice to assign values to variables. It is also
necessary to interpret symbols. Therefore, it is obvious that the propositional
technique for deciding satisfiability does not work in the first-order case.

@@

A position in a term is a tuple of natural numbers, with $\varepsilon$ denoting
the empty tuple. Subterms and positions are inductively defined as follows. Term
$t$ is a subterm of itself at position $\varepsilon$. If $s$ is a subterm of
$t_i$ at position $p$, then $s$ is a subterm of $\cst{f}(\tuplen{t})$ at
position $i.p$. A context is a term with zero or more subterms replaced by a
hole $\square$. We write $C[\overline{u}_n]$ for the term resulting from filling
in the holes of a context $C$ with the terms $\overline{u}_n$, left to right.
We say a term is ground if it has no variables.

Substitutions $\sigma, \varrho, \ldots$ are total mappings from variables to
terms of the same type. Substitutions map only finitely many variables to terms
other than the variable itself. This is denoted as $\{ x_1 \mapsto t_1, \ldots,
x_n \mapsto t_n \}$ where $x_i$ are variables that are not mapped to itself.
Applying a substitution $\sigma$ to a term $t$, denoted $\sigma(t)$ results in
replacing all mapped variables by the corresponding terms.
The composition $\varrho\sigma$ of substitutions is defined by
$\left(\varrho\sigma\right)t=\varrho\left(\sigma t\right)$. We say that substitution is grounding
if it maps all variables to ground terms.

@@

Terms are defined as free variables $F, X, \ldots$, bound variables $x, y, z,
\dotsc$, or symbols $\cst{f}, \cst{g},\allowbreak \cst{a},\allowbreak \cst{b}
\dotsc$. Furthermore, if $s$ and $t$ are terms of type $\tau_1 \rightarrow
\tau_2$, and $\tau_1$, respectively, then $s \, t$ is a term of type $\tau_2$.
Lastly, if $x$ is a bound variable of type $\tau_1$ and $s$ is a term of type
$\tau_2$, then $\lambda$-abstractions $\lamx{s}$ is a term of type $\tau_1
\rightarrow \tau_2$. The syntactic distinction between free and bound variables
gives rise to \emph{loose bound variables} (e.g., $y$ in $\lamx{y \, \cst{a}}$)
\cite{tn-93-patterns}. Out of terms, atoms and formulas are built in the same
way as in first-order logic.

@@

Higher-order substitutions ($\sigma,\varrho,\ldots$) are functions from free and bound
variables to terms. A variable $F$ is mapped by $\sigma$ if $ \sigma F \not\equi
F$. Each substitution maps only finitely many variables, and it is denoted as in
first-order case. Application of  $\sigma$ to term $t$ is denoted $\sigma(t)$;
this application $\alpha$-renames $t$ to avoid variable capture. For example,
$\{X \mapsto x\}(\lamx{X \, a})$ results in $\lam{y}{x \, a}$.  Given a substitution
$\varrho$, which maps $F$ to $s$, we write $\varrho\setminus\{F \mapsto s\}$ to
denote a substitution that does not map $F$ and otherwise coincides with
$\varrho$. Given substitutions $\varrho$ and $\sigma$, which map disjoint sets
of variables, we write $\varrho \cup \sigma$ to denote $\varrho\sigma$.

@@

The resolution and superposition calculi do not work directly on formulas, but
on their simplified forms, clauses. Thus, the initial problem, expressed as a
set of formulas, must be transformed in a set of clauses. For first-order and
higher-order logic, there exists such a transformation that does not affect
satisfiability of the problem \cite{nw-01-small-cnf}.

A literal $l$ is an equation $s \eq t$ or a disequation $s \noteq t$. A clause is a
finite multiset of literals, interpreted and written disjunctively:\ $l_1 \llor
\cdots \llor l_n$. All free variables occurring in the clause are implicitly
universally quantified. Note that $\llor$ used to denote disjunction in clauses
is different than formula-building $\ior$.

In standard, non-clausal, first-order and higher-order logic, an atom  is either
a predicate atom (term of Boolean type) or an equation. For uniformity, and to
stay close to the implementation, we encode predicate atoms as equations with
$\itrue$. Negative literals are encoded as disequations. For example,
$\cst{even}(x)$ is encoded as $\poslit{\cst{even}(x)}$, and
$\neg\,\cst{even}(x)$ is encoded as $\neglit{\cst{even}(x)}$.

Applying substitution to a literal is reduced to applying it to both sides of
the (dis)equa\-tion and it is extended pointwise to clauses. It is written in postfix notation
as $\sigmacl{l}$ and $\sigmacl{C}$. We say $\sigmacl{C}$ is a ground instance of $C$ is $\sigma$ is a grounding
substitution.

@@

Superposition calculus is parametrized by a reduction ordering $\succ$, an
ordering on terms that has the following properties:

@@

The term order is lifted to literals and clauses using the multiset extension of
$\succ$. The order is extended to multisets as follows
\cite[Sect.~2.5]{bg-01-resolution}. We write $S_1 \succ S_2$ if $S_1 \not= S_2$
and whenever $S_2(x) > S_1(x)$ then there is some $y \succ x$ such that $S_1(y)
> S_2(y)$. We use notation $S(x)$ to denote the number of occurrences of $x$ in
$S$. To use multiset extension, positive literals are represented as $ \{ \{s\},
\{t\} \}$ and negative ones as $\{ \{ s, t \} \}$. Clauses are then represented as 
multisets of such literals.

@@

\newcommand{\selfun}{\ensuremath{\mathit{Sel}}}
Literal selection function is a function that selects a (multi)subset of literals.
Superposition requires that at least one of the selected literals is negative.

@@

Calculi from the resolution family, including superposition, perform inferences
only on unifiable terms. We say terms $s$ and $t$ are unifiable if there is a
substitution $\sigma$ such that $\sigmaterm{s} = \sigmaterm{t}$; we further say
$\sigma$ is the unifier. A \emph{unification constraint} $s \unif t$ is an
unordered pair of two terms of the same type. The most general unifier is a
unifier $\sigma$, such that for any other unifier $\theta$, there is a
substitution $\varrho$ such that $\theta = \varrho\sigma$. When there is such a
$\varrho$, we say $\sigma$ is more general than $\theta$. In first-order logic,
the most general unifier is unique up to variable renaming. Furthermore,
first-order logic admits efficient algorithms for computing the most general
unifier \cite{hv-09-unifalgs}.

In higher-order logic, unification is performed modulo rules of
$\alpha\beta\eta$ conversion. Under these conditions, the uniqueness of the most
general unifier is not guaranteed. Consider the constraint $X \,
(\cst{f} \, \cst{a}) \unif \cst{f} \, (X \, \cst{a})$. Any substitution of the
form $X \mapsto \lamx{\cst{f}^i \, x}$ is a unifier, with $\cst{f}^i \, x$
denoting iterated, $i$-fold application of $\cst{f}$. To generalize most general
unifiers to higher-order logic, the concept of complete sets of unifiers was
introduced.

A (higher-order) \emph{unifier} of a multiset of unification constraints $E$ is
a substitution $\sigma$, such that $\sigma s \equi \sigma t$, for all $s \unif t
\in E$. A \emph{complete set of unifiers} (\emph{CSU}) of $E$ is defined as a
set $U$ of $E$'s unifiers along with a set $V$ of \emph{auxiliary variables}
such that no $s \unif t \in E$ contains variables from $V$ and for every unifier
$\rho$ of $E$, there exists a $\sigma \in U$ and a substitution $\theta$ such
that for all $X\not\in V,$ $\rho X = \theta\sigma X$. The most general unifier
corresponds is a one-element CSU. A unifier of terms $s$ and $t$ is a unifier of
the singleton multiset $\{ s \unif t \}$. There is no algorithm to decide
if two higher-order terms are unifiable, but there exist algorithms
that enumerate all elements of CSU for unifiable terms.

These higher-order concepts gracefully generalize the corresponding first-order
ones. For example, if $s$ at $t$ are first-order there must exist a singleton
CSU, i.e., the most general unifier. Thus, we will not explicitly qualify
unifiers as first-order or higher-order and let the context provide enough
information.

@@

Before we spell out the inference rules of standard, first-order superposition
let us introduce some helpful notation, following Schulz \cite{ss-02-brainiac}.
With $t|_p$ we denote subterm of $t$ at position $p$ and with $t[p \rightarrow
t']$ replacement of this subterm with $t'$. Let \selfun be a selection function
and $\succ$ a reduction ordering, $C = l \llor C'$ a clause, and $\sigma$ a
substitution. We say that \sigmacl{l} is eligible for resolution if: (1) nothing
is selected by \selfun and \sigmacl{l} is $\succ$-maximal within literals in
\sigmacl{C} or (2) \selfun selected some literals and \sigmacl{l} is the maximal
within either positive or negative selected literals. We say \sigmacl{l} is
eligible for paramodulation if it is positive, nothing is selected and
\sigmacl{l} is the maximal literal within the literals of \sigmacl{C}. With
$\mgu(s,t)$ we denote the most general unifier of $s$ and $t$. Using horizontal
line to separate premises and the conclusion, four inference rules of
superposition are as follows:

\[
\namedinference{ER}{s \noteq t \llor C}{\sigmacl{C}}
\]
where $\sigma = \mgu(s,t)$, and $(s \noteq t)\sigma$ is eligible
for resolution; \infname{ER} stands for equality resolution
\medskip

\[
\namedinference{EF}{s \eq t \llor u \eq v \llor C}{\sigmacl{(t \noteq v \llor u \eq v \llor C)}}
\]
where $\sigma = \mgu(u,v), \sigma(t) \not\succ \sigma(s)$ and 
$(s \eq t)\sigma$ is eligible for paramodulation; \infname{EF} stands for equality factoring
\medskip

\[
\namedinference{SP}{s \eq t \llor C \qquad u \eq v \llor D}{(u[p \leftarrow t] \eq v \llor C \llor D)\sigma}
\]
where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
$(s \eq t)\sigma$ is eligible for paramodulation, just like $(u \eq v)\sigma$ and $u|_p$ is not a variable;
\infname{SP} stands for positive superposition 
\medskip

@@

Above described rules are used to generate new clauses from already derived
ones. Even though orderings and selection are used to reduce the search space,
it still grows rather fast. To identify unnecessary clauses from the search
space, superposition features \emph{redundancy criterion}
\cite[Sect.~4.2.2]{bg-01-resolution}. We say that a ground clause $C$ is
redundant in a ground set of clauses $N$ if it is entailed by a subset of
clauses in $N$ such that each clause in this subset is smaller than $C$. More
generally, a clause (ground or not) is redundant in set $N$ if every ground
instance of $C$ is redundant for some grounding of $N$. Removing redundant
clauses does not affect completeness of superposition.

@@

\[
\namedinference{RN}{s \eq t \qquad u \noteq v \llor D}{(u[p \leftarrow \sigma(t)] \eq v \llor C \llor D)\sigma}
\]
where $\sigma = \mgu(u|_p, s)$, and $\sigma(s) \succ \sigma(t)$.

Simple rules such as removing duplicate literals, literals of the form $s \noteq
t$ or tautological clauses are clearly justified by the redundancy criterion.
Schulz gives an extensive list of such rules\cite{ss-02-brainiac}.

>>>>

\emph{Atomic formulas} of propositional logic are propositional variables
$\cst{p}, \cst{q}, \ldots$ and constants $\itrue$ and $\ifalse$. More complex
formulas are built inductively using logical connectives $\inot, \iand, \ior,
\iimplies, \allowbreak \iequiv$: if $\phi$ and $\psi$ are formulas, then $\inot \phi, \phi \iand
\psi, \allowbreak \phi \ior \psi, \phi \iimplies \psi, \phi \iequiv \psi$ are formulas as well.

To interpret the formulas, propositional variables are assigned 0 (false) or 1
(true), constants $\itrue$ and $\ifalse$ are assigned 1 and 0, respectively; the
formula is interpreted using the rules for each connective
\cite[Sect.~1.4]{hr-00-logic-in-cs}. As there are finitely many propositional
variables in a formula, trying all (finitely many) possible variable assignments
describes an algorithm to decide satisfiability (existence of satisfiable
assignment for a formula) or validity (if all assignments are satisfy the formula).

@@

We distinguish a set of base types $T$ which is required to have the Boolean
type $o$, and a set of symbols $\Sigma$. To each symbol $\cst{f} \in \Sigma$ a
tuple $(\tau_1, \ldots, \tau_{n}, \tau), n \geq 0$ of types is assigned, written
$\cst{f} : (\tau_1, \ldots, \tau_{n}) \rightarrow \tau$. We say that $(\tau_1,
\ldots, \tau_{n})$ are \emph{argument types}, $\tau$ is the \emph{return type}, and that $n$
is the \emph{arity} of the symbol $\cst{f}$. If return type of $\cst{f}$ is $o$ we call
$\cst{f}$ a \emph{predicate symbol}, otherwise we call it a \emph{function symbol}. Argument types may not be Boolean.

\emph{Terms} are the basic building blocks of first-order logic. Variables
$x,y,\ldots$, assigned types $\tau \in T$, are terms. If $t_1,\ldots,t_n$ are
terms of types $\tau_1, \ldots, \tau_n$, respectively, and $\cst{f} : (\tau_1,
\ldots, \tau_{n}) \rightarrow \tau$, then $\cst{f}(t_1, \ldots, t_n)$ is a term
of type $\tau$. If $n=0$, we drop parentheses and write $\cst{f}$. We also
abbreviate $\cst{f}(t_1, \ldots, t_n)$ to $\cst{f}(\tuplen{t})$. Using similar
notation, we abbreviate a tuple of terms $(t_1, \ldots, t_n)$ as $(\tuplen{t})$
or simply $\tuplen{t}$.

Terms are used to build \emph{atoms}. Atom is either a term $t$ of type $o$ (\emph{predicate atom}) or an
equation $s \eq t$ where terms $t$ and and $s$ are of the same type (\emph{equational atom}). Atoms are
combined using logical connectives to build formulas just like in propositional
logic. Additionally, first-order formulas are built using quantifiers: if $\phi$
is a formula and $x$ is a variable then $\iforall x.\, \phi$, as well as
$\iexists x.\, \phi$ are formulas. The first formula requires that $\phi$ holds
for any value of $x$, while the second one requires that there is some $x$ for
which $\phi$ holds. As soon as there is a single functional symbol with arity
greater than 0, there are infinitely many terms that can be substituted for a
free variable. Furthermore, in first-order logic, it does not suffice to assign
values to variables to determine the truth value of the formula. It is also
necessary to interpret symbols. Therefore, it is obvious that the propositional
technique for deciding satisfiability does not work in the first-order case.

@@

A {\em position} in a term is a tuple of natural numbers, with $\varepsilon$ denoting
the empty tuple. {\em Subterms} and positions are inductively defined as follows. Term
$t$ is a subterm of itself at position $\varepsilon$. If $s$ is a subterm of
$t_i$ at position $p$, then $s$ is a subterm of $\cst{f}(\tuplen{t})$ at
position $i.p$. A {\em context} is a term with zero or more subterms replaced by a
hole $\square$. We write $C[\overline{u}_n]$ for the term resulting from filling
in the holes of a context $C$ with the terms $\overline{u}_n$, left to right.
We say a term is {\em ground} if it has no variables.

{\em Substitutions} $\sigma, \varrho, \ldots$ are total mappings from variables to
terms of the same type. Substitutions map only finitely many variables to terms
other than the variable itself. This is denoted as $\{ x_1 \mapsto t_1, \ldots,
x_n \mapsto t_n \}$ where $x_i$ are variables that are not mapped to itself.
Applying a substitution $\sigma$ to a term $t$, denoted $\sigma(t)$ results in
replacing all mapped variables by the corresponding terms.
The composition $\varrho\sigma$ of substitutions is defined by
$\left(\varrho\sigma\right)t=\varrho\left(\sigma t\right)$. We say that substitution is {\em grounding}
if it maps all variables to ground terms.

@@

Terms are defined as free variables $F, X, \ldots$, bound variables $x, y, z,
\dotsc$, or symbols $\cst{f}, \cst{g},\allowbreak \cst{a},\allowbreak \cst{b}
\dotsc$. Furthermore, if $s$ and $t$ are terms of type $\tau_1 \rightarrow
\tau_2$, and $\tau_1$, respectively, then $s \, t$ is a term of type $\tau_2$.
Lastly, if $x$ is a bound variable of type $\tau_1$ and $s$ is a term of type
$\tau_2$, then $\lambda$-abstraction $\lamx{s}$ is a term of type $\tau_1
\rightarrow \tau_2$. The syntactic distinction between free and bound variables
gives rise to \emph{loose bound variables} (e.g., $y$ in $\lamx{y \, \cst{a}}$)
\cite{tn-93-patterns}. Out of terms, atoms and formulas are built in the same
way as in first-order logic.

@@

Higher-order substitutions ($\sigma,\varrho,\ldots$) are functions from free and bound
variables to terms. A variable $F$ is mapped by $\sigma$ if $ \sigma(F) \not\equi
F$. Each substitution maps only finitely many variables, and it is denoted as in
the first-order case. Application of  $\sigma$ to term $t$ is denoted $\sigma(t)$;
this application $\alpha$-renames $t$ to avoid variable capture. For example,
$\{X \mapsto x\}(\lamx{X \, \cst{a}})$ results in $\lam{y}{x \, \cst{a}}$.  Given a substitution
$\varrho$, which maps $F$ to $s$, we write $\varrho\setminus\{F \mapsto s\}$ to
denote a substitution that does not map $F$ and otherwise coincides with
$\varrho$. Given substitutions $\varrho$ and $\sigma$, mapping disjoint variable
sets, we write $\varrho \cup \sigma$ to denote $\varrho\sigma$.

@@

The resolution and superposition calculi do not work directly on formulas, but
on their simplified forms, {\em clauses}. Thus, the initial problem, expressed as a
set of formulas, must be transformed in a set of clauses. For all three
discussed logic, there exists such a transformation that does not affect
satisfiability of the problem \cite{nw-01-small-cnf}.

A {\em literal} $l$ is an equation $s \eq t$ or a disequation $s \noteq t$. A clause is a
finite multiset of literals, interpreted and written disjunctively:\ $l_1 \llor
\cdots \llor l_n$. All free variables occurring in the clause are implicitly
universally quantified. Note that $\llor$ used to denote disjunction in clauses
is different than formula-building $\ior$.

In standard, non-clausal, first-order and higher-order logic, an atom is either
predicate or equational. For uniformity, and to
stay close to the implementation, we encode predicate atoms as equations with
$\itrue$. Negative literals are encoded as disequations. For example,
$\cst{even}(x)$ is encoded as $\poslit{\cst{even}(x)}$, and
$\neg\,\cst{even}(x)$ is encoded as $\neglit{\cst{even}(x)}$.

Applying substitution to a literal is reduced to applying it to both sides of
the (dis)equa\-tion and it is extended pointwise to clauses. It is written in
postfix notation as $\sigmacl{l}$ and $\sigmacl{C}$. We say $\sigmacl{C}$ is a
ground instance of $C$ if $\sigma$ is a grounding substitution.

@@

Superposition calculus is parametrized by a {\em reduction ordering} $\succ$, an
ordering on terms that has the following properties:

@@

The term order is lifted to literals and clauses using the {\em multiset extension} of
$\succ$. The order is extended to multisets as follows
\cite[Sect.~2.5]{bg-01-resolution}. For two multisets $S_1$ and $S_2$, we write $S_1 \succ S_2$ if $S_1 \not= S_2$
and whenever $S_2(x) > S_1(x)$ then there is some $y \succ x$ such that $S_1(y)
> S_2(y)$. We use notation $S(x)$ to denote the number of occurrences of $x$ in
$S$. To use multiset extension, positive literals are represented as $ \{ \{s\},
\{t\} \}$ and negative ones as $\{ \{ s, t \} \}$. Clauses are then represented as 
multisets of such literals.

@@

\newcommand{\selfun}{\ensuremath{\mathit{Sel}}}
{\em Literal selection function} is a function that selects a (multi)subset of literals.
Superposition requires that at least one of the selected literals is negative.

@@

Calculi from the resolution family, including superposition, perform inferences
only on unifiable terms. We say terms $s$ and $t$ are {\em unifiable} if there is a
substitution $\sigma$ such that $\sigmaterm{s} = \sigmaterm{t}$; we further say
$\sigma$ is the unifier. A \emph{unification constraint} $s \unif t$ is an
unordered pair of two terms of the same type. A \emph{most general unifier} is a
unifier $\sigma$, such that for any other unifier $\theta$, there is a
substitution $\varrho$ such that $\theta = \varrho\sigma$. When there is such a
$\varrho$, we say $\sigma$ is more general than $\theta$. In first-order logic,
the most general unifier is unique up to variable renaming. Furthermore,
first-order logic admits efficient algorithms for computing the most general
unifier \cite{hv-09-unifalgs}.

In higher-order logic, unification is performed modulo rules of
$\alpha\beta\eta$ conversion. Under these conditions, the uniqueness of the most
general unifier is not guaranteed. Consider the constraint $X \,
(\cst{f} \, \cst{a}) \unif \cst{f} \, (X \, \cst{a})$. Any substitution of the
form $\{ X \mapsto \lamx{\cst{f}^i \, x} \}$ is a unifier, with $\cst{f}^i \, x$
denoting iterated, $i$-fold application of $\cst{f}$. To generalize most general
unifiers to higher-order logic, the concept of {\em complete sets of unifiers} was
introduced.

A (higher-order) unifier of a multiset of unification constraints $E$ is
a substitution $\sigma$, such that $\sigma s \equi \sigma t$, for all $s \unif t
\in E$. A complete set of unifiers (\emph{CSU}) of $E$ is defined as a
set $U$ of $E$'s unifiers along with a set $V$ of \emph{auxiliary variables}
such that no $s \unif t \in E$ contains variables from $V$ and for every unifier
$\rho$ of $E$, there exists a $\sigma \in U$ and a substitution $\theta$ such
that for all $X\not\in V,$ $\rho X = \theta\sigma X$. The most general unifier
corresponds is a one-element CSU. A unifier of terms $s$ and $t$ is a unifier of
the singleton multiset $\{ s \unif t \}$. There is no algorithm to decide
if two higher-order terms are unifiable, but there exist algorithms
that enumerate all elements of CSU for unifiable terms.

These higher-order concepts gracefully generalize the corresponding first-order
ones. For example, if $s$ at $t$ are first-order there must exist a singleton
CSU, i.e., the most general unifier. Thus, we will not explicitly qualify
unifiers as first-order or higher-order and ensure the context provides enough
information.

@@

Before we spell out the inference rules of standard, first-order superposition
let us introduce some helpful notation, following Schulz \cite{ss-02-brainiac}.
With $t|_p$ we denote subterm of $t$ at position $p$ and with $t[p \leftarrow
t']$ replacement of this subterm with $t'$. Let \selfun{} be a selection function,
$\succ$ a reduction ordering, $C = l \llor C'$ a clause, and $\sigma$ a
substitution. We say that \sigmacl{l} is eligible for resolution if: (1) nothing
is selected by \selfun{} and \sigmacl{l} is $\succ$-maximal within literals in
\sigmacl{C} or (2) \selfun{} selected some literals and \sigmacl{l} is the maximal
within either positive or negative selected literals. We say \sigmacl{l} is
eligible for paramodulation if it is positive, nothing is selected and
\sigmacl{l} is the maximal literal within the literals of \sigmacl{C}. With
$\mgu(s,t)$ we denote the most general unifier of $s$ and $t$. Using horizontal
line to separate premises and the conclusion, four inference rules of
superposition are as follows:

\[
\namedinference{ER}{s \noteq t \llor C}{\sigmacl{C}}
\]
where $\sigma = \mgu(s,t)$, and $(s \noteq t)\sigma$ is eligible
for resolution; \infname{ER} stands for equality resolution

\[
\namedinference{EF}{s \eq t \llor u \eq v \llor C}{\sigmacl{(t \noteq v \llor u \eq v \llor C)}}
\]
where $\sigma = \mgu(s,u), \sigma(t) \not\succ \sigma(s)$ and 
$(s \eq t)\sigma$ is eligible for paramodulation; \infname{EF} stands for equality factoring

\[
\namedinference{SP}{s \eq t \llor C \qquad u \eq v \llor D}{(u[p \leftarrow t] \eq v \llor C \llor D)\sigma}
\]
where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
$(s \eq t)\sigma$ is eligible for paramodulation, just like $(u \eq v)\sigma$ and $u|_p$ is not a variable;
\infname{SP} stands for positive superposition

@@

Above described rules are used to generate new clauses from already derived
ones. Even though orderings and selection are used to reduce the search space,
it still grows rather fast. To identify unnecessary clauses in the search
space, superposition features \emph{redundancy criterion}
\cite[Sect.~4.2.2]{bg-01-resolution}. We say that a ground clause $C$ is
{\em redundant} in a ground set of clauses $N$ if it is entailed by a subset of
clauses in $N$ such that each clause in this subset is smaller than $C$. More
generally, a clause (ground or not) is redundant in set $N$ if every ground
instance of $C$ is redundant for some grounding of $N$. Removing redundant
clauses does not affect completeness of superposition.

@@

\[
\namedsimp{RN}{s \eq t \qquad u \noteq v \llor D}
              {s \eq t \qquad (u[p \leftarrow \sigma(t)] \eq v \llor C \llor D)\sigma}
\]
where $\sigma = \mgu(u|_p, s)$, and $\sigma(s) \succ \sigma(t)$.

Simple rules such as removing duplicate literals, literals of the form $s \noteq
s$ or tautological clauses are clearly justified by the redundancy criterion.
Schulz gives an extensive list of such rules \cite{ss-02-brainiac}.
